State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-13 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}\left(\mathbf{r}\right)}{1 + \exp \left( \varsigma \left( v_0\left( \mathbf{r} \right) - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameters $f_{max}$ and $v_0$ describe the maximum firing rate and firing threshold of the neural populations. The parameter $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into \ref{SpikesToPotential} we get the spatiotemporal model 
\begin{eqnarray}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) &=&  \int_{-\infty}^t h\left(t - t'\right) \\
	&\times&\int_\Omega w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}'dt'. \nonumber
\end{eqnarray}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} Dh\left( t \right) = \delta \left( t \right), 
\end{equation}
where $D=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{eqnarray}
	\label{FinalFormContinuous} \frac{dv\left( \mathbf{r},t \right)}{dt} &+& \zeta v\left( \mathbf{r},t \right) = \\
	&&\int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'}. \nonumber
\end{eqnarray}
To arrive at the integro-difference equation (IDE) form of the model we discretise time using a first-order Euler method (see~\ref{Time Discretization}) giving 
\begin{eqnarray}
	\label{DiscreteTimeModel} v_{t+1}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 
	&+& e_t\left(\mathbf{r}\right), 
\end{eqnarray}
where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is i.i.d. such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$ denotes a spatial Gaussian Process with mean zero and covariance function $\gamma(\mathbf{s}-\mathbf{r})$ \cite{Rasmussen2006} This term was added to account for uncertainty in the model. The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors, $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ and $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate Normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

Let the spectral representation of the dynamic field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. Given $V_t(\boldsymbol{\nu})$, the distance between adjacent sensors should be no more than
% 
% Then, for an accurate representation, $V_t(\boldsymbol{\nu})$ needs to be spatially band-limited. Such a condition is usually satisfied due to the band-limiting effect of most practical kernels and covariance functions \todo{~cite\{some EEG literature here\}} \omg{will try to find something. These kernel are kinda theoretical and I have not ever seen and estimate of the covariance function. Can we remove the sentence ``Such a condition is usually satisfied...'' }. Thus we can assume that the spectral response $V_t(\boldsymbol{\nu})$ is band-limited with $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$,  
% 
% Note that for the two-dimensional, homogeneous and isotropic fields being considered $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field the distance between adjacent sensors, $\Delta_s$, needs to satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_s < \frac{1}{2\rho_s\nu_{c}}, 
\end{equation}
where $\boldsymbol{\nu}_c$ the cut-off frequency, $\rho_s \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied for the hidden dynamic field $v_t(\mathbf{r})$ to be precisely inferred from the sampled observations, $\mathbf{y}_t$. If the condition is not satisfied, spatial aliasing will occur.

@@

\begin{algorithm}
	\begin{small}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
		 \hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propogate them through equation \ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
		\begin{equation*}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f,u_t) 
		\end{equation*}
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman Filter update equations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation*}
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation*}
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propogate them through equation \ref{eq:QmatrixForSigmapoints}
		\begin{equation*}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b,u_t) 
		\end{equation*}
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		\end{equation*}
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{small}
\end{algorithm}

@@

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the parameter and state estimation, $250$ realisations of the data were generated. Each realisation consisted of 200 ms of data and the estimation was applied to the final 200 ms, allowing the model to stabilise. The initial state and parameters were unknown to the estimator. The distribution of the parameters and the resulting kernel are shown in figure~\ref{fig:Parameters} and figure~\ref{fig:KernelEstimates} respectively. Although a wide distribution can be seen over the parameter estimates the ratio between them remans in a narrow range as shown in figure~\ref{fig:ParametersRatio}. The state estimates for one realization were evaluated by comparing the field reconstruction to the true field using the mean over time of the Root Mean Square Error (RMSE) and standard deviation of the RMSE giving $0.19\pm 0.01$. The RMSE of the field estimate over time is also shown in figure~\ref{fig:RMSE}. One snapshot for the original and the field estimate and their corresponding error are illustrated in figure~\ref{fig:FieldEstimate}.
\begin{figure*}
\subfigure[][]{\includegraphics{./Graph/KernelEstimate_skinny.pdf}}
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_skinny.pdf}}\\
\subfigure[][]{\includegraphics{./Graph/KernelEstimate_wide.pdf}}
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_wide.pdf}}\\
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_skinny_db.pdf}}

>>>>

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameters $f_{max}$ and $v_0$ describe the maximum firing rate and firing threshold of the neural populations. The parameter $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into \ref{SpikesToPotential} we get the spatiotemporal model 
\begin{eqnarray}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) &=&  \int_{-\infty}^t h\left(t - t'\right) \\
	&\times&\int_\Omega w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}'dt'. \nonumber
\end{eqnarray}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} Dh\left( t \right) = \delta \left( t \right), 
\end{equation}
where $D=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{eqnarray}
	\label{FinalFormContinuous} \frac{dv\left( \mathbf{r},t \right)}{dt} &+& \zeta v\left( \mathbf{r},t \right) = \\
	&&\int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'}. \nonumber
\end{eqnarray}
To arrive at the integro-difference equation (IDE) form of the model we discretise time using a first-order Euler method (see~\ref{Time Discretization}) giving 
\begin{eqnarray}
	\label{DiscreteTimeModel} v_{t+T_s}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 
	&+& e_t\left(\mathbf{r}\right), 
\end{eqnarray}
where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is i.i.d. such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$ denotes a spatial Gaussian Process with mean zero and covariance function $\gamma(\mathbf{s}-\mathbf{r})$ \cite{Rasmussen2006} This term was added to account for uncertainty in the model. The mapping between the membrane voltage and the collected iEEG/LFP data is modelled using the observation function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field, $n=1,...,N$ indexes the sensors, $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ and $\mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$ denotes the multivariate Normal distribution with mean zero and covariance matrix $\Sigma_{\varepsilon}$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the sensor pick-up geometry where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

Let the spectral representation of the dynamic field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$, where $\nu$ denotes spatial frequency. Given $V_t(\boldsymbol{\nu})$, the distance between adjacent sensors should be no more than
% 
% Then, for an accurate representation, $V_t(\boldsymbol{\nu})$ needs to be spatially band-limited. Such a condition is usually satisfied due to the band-limiting effect of most practical kernels and covariance functions \todo{~cite\{some EEG literature here\}} \omg{will try to find something. These kernel are kinda theoretical and I have not ever seen and estimate of the covariance function. Can we remove the sentence ``Such a condition is usually satisfied...'' }. Thus we can assume that the spectral response $V_t(\boldsymbol{\nu})$ is band-limited with $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$,  
% 
% Note that for the two-dimensional, homogeneous and isotropic fields being considered $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field the distance between adjacent sensors, $\Delta_s$, needs to satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_s < \frac{1}{2\rho_s\nu_{c}}, 
\end{equation}
where $\boldsymbol{\nu}_c$ the cut-off frequency, $\rho_s \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied for the hidden dynamic field $v_t(\mathbf{r})$ to be precisely inferred from the sampled observations, $\mathbf{y}_t$. If the condition is not satisfied, spatial aliasing will occur.

@@

\begin{algorithm}
	\begin{small}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
		 \hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
		\begin{equation*}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) 
		\end{equation*}
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman Filter update equations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation*}
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation*}
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propogate them through equation \ref{eq:QmatrixForSigmapoints}
		\begin{equation*}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b,u_t) 
		\end{equation*}
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		\end{equation*}
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{small}
\end{algorithm}

@@

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the parameter and state estimation, $250$ realisations of the data were generated. Each realisation consisted of 200 ms of data and the estimation was applied to the final 195 ms, allowing the model to stabilise.\omg{transient!} The initial state and parameters were unknown to the estimator. The distribution of the parameters and the resulting kernel are shown in figure~\ref{fig:Parameters} and figure~\ref{fig:KernelEstimates} respectively. Although a wide distribution can be seen over the parameter estimates the ratio between them remans in a narrow range as shown in figure~\ref{fig:ParametersRatio}. The state estimates for one realisation were evaluated by comparing the field reconstruction to the true field using the mean over time of the Root Mean Square Error (RMSE) and standard deviation of the RMSE giving $0.19\pm 0.01$. The RMSE of the field estimate over time is also shown in figure~\ref{fig:RMSE}. One snapshot for the original and the field estimate and their corresponding error are illustrated in figure~\ref{fig:FieldEstimate}.

\begin{figure*}
\subfigure[][]{\includegraphics{./Graph/KernelEstimate_skinny.pdf}}
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_skinny.pdf}}\\
\subfigure[][]{\includegraphics{./Graph/KernelEstimate_wide.pdf}}
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_wide.pdf}}\\
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT_skinny_db.pdf}}
